An Extendable \proglang{Python} Implementation of Robust Optimisation Monte Carlo

### 2022-12-28 article/article.tex
finished chapter 4

In this paper, we present the implementation of ROMC at the
\proglang{Python} package \pkg{ELFI} (Engine for likelihood-free
inference)~\citet{1708.00707}. The implementation has been designed to
ensure extensibility. ROMC is an LFI framework; it defines a sequence
of algorithmic steps for aproximating the posterior without enforcing
a specific algorithm for solving each step. Therefore, a researcher
may use ROMC as the backbone algorithm and develop novel methods to
solve each separate step.\footnote{For being a ready-to-use
  algorithm,~\citet{Ikonomov2019} proposed a default method for each
  step, but this choice is by no means restrictive.} We have designed
our software for facilitating such experimentation.

@@

%% -- Design Principles -----------------------------------------------------
\section{Implementation}
\label{sec:implementation}
This section is split in two parts. We first express ROMC as an
algorithm and then we present the general implementation principles we
follow.

\subsection{Algorithmic view of ROMC}
\input{latex_files/ch3/3-0-design-principles.tex}

\subsection{General implementation principles}
\label{subsec:general_design}
\input{latex_files/ch3/3-1-implementation-principles.tex}

%% -- ROMC implementation --
\section{Implemented functionalities}
\input{latex_files/ch4/4-0-intro.tex}

\subsection{Training part}
\label{subsec:training}
\input{latex_files/ch4/4-1-training.tex}

\subsection{Inference part}
\label{subsec:inference}
\input{latex_files/ch4/4-2-inference.tex}

\subsection{Evaluation part}
\label{subsec:evaluation}
\input{latex_files/ch4/4-3-evaluation.tex}

\subsection{Extensibility} % better title
\label{subsec:extensibility}
\input{latex_files/ch4/4-4-extensibility.tex}

>>>>

In this paper, we present the implementation of ROMC at the
\proglang{Python} package \pkg{ELFI} (Engine for likelihood-free
inference)~\citet{1708.00707}. The implementation has been designed to
ensure extensibility. ROMC is an LFI framework; it defines a sequence
of algorithmic steps for approximating the posterior without enforcing
a specific algorithm for solving each step. Therefore, a researcher
may use ROMC as the backbone algorithm and develop novel methods to
solve each separate step.\footnote{For being a ready-to-use
  algorithm,~\citet{Ikonomov2019} proposed a default method for each
  step, but this choice is by no means restrictive.} We have designed
our software for facilitating such experimentation.

@@

%% -- Design Principles -----------------------------------------------------
\section{Overview of the implementation}
\label{sec:implementation}
In this section, we express ROMC as an algorithm and then we present
the general implementation principles.

\subsection{Algorithmic view of ROMC}

For designing an extendable implementation, we firstly define ROMC as
a sequence of algorithmic steps. At a high level, ROMC can be split
into the training and the inference part; the training part covers the
steps for estimating the proposal regions and the inference part
calculates the weighted samples. In
Algorithm~\ref{alg:romc_algorithm}, that defines ROMC as an algorithm,
steps 2-11 (before the horizontal line) refer to the training part and
steps 13-18 to the inference part.

\begin{algorithm}[!ht]
  \caption{ROMC. Requires the prior \( p(\thetab) \), the simulator
    \(M_r(\thetab)\), number of optimization problems \(n_1\), number
    of samples per region \(n_2\), acceptance limit
    \(\epsilon\)}\label{alg:romc_algorithm}
  \begin{algorithmic}[1]
    \Procedure{ROMC}{}
    \For{\(i \gets 1 \textrm{ to } n_1\)}
    \State \(\ub_i \sim p(\ub)\) \Comment{Draw nuisance variables}
    \State Convert \(M_r(\thetab) \) to \( g(\thetab, \ub=\ub_i) \) \Comment{Define deterministic simulator}
      \State \( d_i(\thetab) = d(g(\thetab, \ub=\ub_i), \yb_0) \) \Comment{Define distance function}
      \State \(\thetab_i^* = \text{argmin}_{\thetab} d_i\), \(d_i^*=d_i(\thetab_i^*)\) \Comment{Solve optimization problem}
      \If{\(d^*_i > \epsilon\)}
        \State Go to 2 \Comment{Filter solution}
      \EndIf
      \State Estimate \(\hat{\accregioni}\) and define \(q_i\) \Comment{Estimate proposal area}
      \State (Optional) Fit \(\Tilde{d}_i\) on \(\hat{\accregioni}\) \Comment{Fit surrogate model}
      \\\hrulefill
      \For{\(j \gets 1 \textrm{ to } n_2\)}
      \State \(\thetab_{ij} \sim q_i\), compute \(w_{ij}\) as in Algorithm \ref{alg:sampling_GB} \Comment{Sample}
      \EndFor
    \EndFor
    \State \(\E_{p(\thetab|\data)}[h(\thetab)]\) as in eq.~\eqref{eq:expectation} \Comment{Estimate an expectation}
    \State \(p_{d,\epsilon}(\thetab) \) as in eq.~\eqref{eq:approx_posterior_omc} \Comment{Evaluate the unnormalized posterior}
    \EndProcedure
  \end{algorithmic}
\end{algorithm}

\subsubsection*{Training part}
\noindent
At the training (fitting) part, the goal is the estimation of the
proposal regions \(\hat{\accregioni}\), which expands to three
mandatory tasks; (a) sample the nuisance variables
\(\ub_i \sim p(\ub)\) for defining the deterministic distance
functions \(d_i(\thetab)\) (Steps 3-5), (b) solve the optimization
problems for obtaining \(\thetab_i^*, d_i^*\) and keep the solutions
inside the threshold \(\epsilon\)(Steps 6-9), and (c) estimate the
bounding boxes \(\accregionihat\) to define uniform distributions
\(q_i\) on them (Step 10). Optionally, a surrogate model
\(\tilde{d}_i\) can be fitted for a faster inference phase.

If \(d_i(\thetab)\) is differentiable, using a gradient-based method
is advised for obtaining \(\thetab_i^*\) faster. In this case, the
gradients \(\nabla_{\thetab} d_i\) gradients are approximated
automatically with finite-differences. This approximation requires two
evaluations of \(d_i\) for each parameter
\(\theta_m, m \in \{1, \ldots, D\}\), which scales well only in
low-dimensional problems. If \(d_i(\thetab)\) is not differentiable,
Bayesian Optimization can be used as an alternative. In this scenario,
the training part becomes slower due to fitting of the surrogate model
and the blind optimization steps.

After obtaining the optimal points \(\thetab^*_i\), we estimate the
proposal regions. Algorithm~\ref{alg:region_construction} describes
the line search approach for finding the region boundaries. The axes
of each bounding box \(\vb_m, m = \{1, \ldots, D\}\) are defined as
the directions with the highest curvature at \(\thetab_i^*\) computed
by the eigenvalues of the Hessian matrix \(\hessian_i\) of \(d_i\) at
\(\thetab_i\) (Step 1). Depending on the algorithm used in the
optimization step, we either use the real distance \(d_i\) or the
Gaussian Process approximation \(\hat{d}_i\). When the distance
function is the Euclidean distance (default choice), the Hessian
matrix can be also computed as \(\hessian_i = \jac_i^T\jac_i\), where
\(\jac_i\) is the Jacobian matrix of the summary statistics
\(\Phi(g(\thetab, \ub_i))\) at \(\thetab_i^*\). This approximation has
the computational advantage of using only first-order
derivatives. After defining the axes, we search for the bounding box
limits with a line step algorithm (Steps 2-21). The key idea is to
take long steps \(\eta\_\text{start}\) until crossing the boundary
where \(d > \epsilon\) and then take small steps backwards to find the
exact boundary position.

\begin{algorithm}[!ht]
  \caption{Approximation \(\accregioni\) with a bounding box \(\hat{\accregioni}\);
  Requires: a model of distance \(d_i(\thetab)\),
  an optimal point \(\thetab_i^*\),
  a number of refinements \(K\),
  a step size \(\eta\_\text{start}\),
  maximum iterations \(M\) and
  a curvature matrix \(\hessian_i\) (\(\jac_i^T\jac_i \) or GP Hessian)}\label{alg:region_construction}
  \begin{algorithmic}[1]
  \State Compute eigenvectors \(\vb_{m}\) of \(\hess_i\) {\scriptsize (\(m = 1,\ldots,D)\)}
  \For{\(m \gets 1 \textrm{ to } D\)}
    \State \(\Tilde{\thetab} \gets \thetab_i^*\) \label{algstep:box_constr_start}
    \State \(k \gets 0\)
    \State \(\eta \gets \eta\_\text{start}\) \Comment{Initialize \(\eta\)}
    \Repeat
          \State \(j \gets 0\)
          \Repeat
            \State \(\Tilde{\thetab} \gets \Tilde{\thetab} + \eta \ \vb_{m}\) \Comment{Large step size \(\eta\).}
            \State \(j \gets j + 1\)
          \Until{\(d(g(\Tilde{\thetab}, \ub=\ub_i), \data) > \epsilon\) or \(j \geq M\)} \Comment{Check distance or maximum iterations}
          \State \(\Tilde{\thetab} \gets \Tilde{\thetab} - \eta \ \vb_{m}\)
          \State \(\eta \gets \eta/2\) \Comment{More accurate region boundary}
          \State \(k \gets k + 1\)
      \Until \(k = K\)
      \If{\(\Tilde{\thetab} = \thetab_i^*\)} \Comment{Check if no step has been done}
        \State \(\Tilde{\thetab} \gets \Tilde{\thetab} + \dfrac{\eta\_{\text{start}}}{2^K} \vb_m\) \Comment{Then, make the minimum step}
      \EndIf \label{algstep:box_constr_end}
      \State Set \(\Tilde{\thetab}\) as the positive end point along \(\vb_m\)
      \State Run steps~\ref{algstep:box_constr_start}~-~\ref{algstep:box_constr_end} for \(\vb_m = - \vb_m\) and set \(\Tilde{\thetab}\) as the negative end point along \(\mathbf{v}_{m}\)
  \EndFor
  \State Fit a rectangular box around the region end points and define \(q_i\) as uniform distribution
  \end{algorithmic}
\end{algorithm}

\subsubsection*{Inference Part}

The inference part includes one or more of the following three tasks;
(a) sample from the posterior distribution
\( \thetab_i \sim p_{d, \epsilon}(\thetab|\data)\)
(Equation~\ref{eq:sampling}), (b) compute an expectation
\(\E_{\thetab|\data}[h(\thetab)]\) (Equation~\ref{eq:expectation})
and/or (c) evaluate the unnormalized posterior
\(p_{d, \epsilon}(\thetab|\data)\)
(Equation~\ref{eq:approx_posterior_omc}). Sampling is performed by
getting \(n_2\) samples from each proposal distribution \(q_i\). For
each sample \(\thetab_{ij}\), the distance function\footnote{As
  before, if a surrogate model \(\hat{d}\) is available, it can be
  utilized as the distance function.} is evaluated for checking if it
lies inside the acceptance region. Algorithm~\ref{alg:sampling_GB}
defines the steps for computing a weighted sample. After we obtain
weighted samples, computing the expectation is straightforward using
Equation~\ref{eq:expectation}. Finally, evaluating the unnormalized
posterior at a specific point \(\thetab\) requires access to the
distance functions \(d_i\) and the prior distribution
\(p(\thetab)\). Following Equation~\ref{eq:approx_posterior_omc}, the
evaluation involves counting how many distance functions rating over
all \(d_i(\thetab) < \epsilon\). It is worth noticing that for
evaluating the unnormalized posterior, there is no need for solving
the optimization problems and building the proposal regions.

\begin{algorithm}[H]
    \centering
    \caption{Sampling. Requires a function of distance \(d_i\), the prior distribution \(p(\thetab)\), the proposal distribution \(q_i\)}\label{alg:sampling_GB}
    \begin{algorithmic}[1]
      \State \(\thetab_{ij} \sim q_i\)
          \If {\(d_i(\thetab_{ij}) > \epsilon\)}
            \State Go to 2 \Comment{Reject sample}
          \Else {}
            \State \(w_{ij} = \frac{p(\thetab_{ij})}{q(\thetab_{ij})}\) \Comment{Compute weight}
            \State Store \((w_{ij}, \thetab_{ij})\) \Comment{Store weighted sample}
          \EndIf
    \end{algorithmic}
\end{algorithm}

\subsection{General implementation principles}
\label{subsec:general_design}

\input{latex_files/graphs/romc_graph}

The overview of our implementation is illustrated in
Figure~\ref{fig:romc_overview}. Following \proglang{Python}'s naming
principles, the methods starting with an underscore (green rectangles)
represent internal (private) functions, whereas the rest (blue
rectangles) are the methods exposed at the API. In
Figure~\ref{fig:romc_overview}, it can be easily observed that the
implementation follows Algorithm~\ref{alg:romc_algorithm}. The
training part includes all the steps until the computation of the
proposal regions, i.e., sampling the nuisance variables, defining the
optimization problems, solving them, constructing the regions and
fitting local surrogate models. The inference part comprises of
evaluating the unnormalized posterior (and the normalized when is
possible), sampling and computing an expectation. We also provide some
utilities for inspecting the training process, such as plotting the
histogram of the final distances or visualizing the constructed
bounding boxes. Finally, in the evaluation part, we provide two
methods for evaluating the inference; (a) computing the Effective
Sample Size (ESS) of the samples and (b) measuring the divergence
between the approximate posterior the ground-truth, if the latter is
available.\footnote{Normally, the ground-truth posterior is not
  available; However, this functionality is useful in cases where the
  posterior can be computed numerically or with an alternative method,
  e.g., Rejection Sampling, and we want to measure the discrepancy
  between the two approximations.}

\subsubsection*{Parallelizing ROMC method}

As discussed, ROMC has the significant advantage of being fully
parallelisable. We exploit this fact by implementing a parallel
version of the major fitting components; (a) solving the optimization
problems, (b) constructing bounding box regions. We parallelize these
processes using the built-in \proglang{Python} package
\pkg{multiprocessing}. The specific package enables concurrency, using
sub-processes instead of threads, for side-stepping the Global
Interpreter (GIL). For activating the parallel version of the
algorithm, the user simply has to set \code{elfi.ROMC(<output_node>,
  parallelize = True)}.

\subsubsection*{Simple one-dimensional example}

For illustrating the functionalities we will use the following running
example introduced by \citet{Ikonomov2019},

\begin{gather} \label{eq:1D_example}
  p(\theta) = \mathcal{U}(\theta;-2.5,2.5)\\ \label{eq:1D_example_eq_2}
  p(y|\theta) = \left\{
    \begin{array}{ll} \theta^4 + u & \mbox{if } \theta \in [-0.5, 0.5]
\\ |\theta| - c + u & \mbox{otherwise}
    \end{array} \right.\\
  u \sim \mathcal{N}(0,1)
\end{gather}

\noindent

The prior is a uniform distribution in the range \([-2.5, 2.5]\) and
the likelihood is defined at Equation~\ref{eq:1D_example_eq_2}. The
constant \(c\) is \(0.5 - 0.5^4\) ensures that the PDF is
continuous. There is only one observation \(y_0 = 0\). The inference
in this particular example can be performed quite easily without using
a likelihood-free inference approach. We can exploit this fact for
validating the accuracy of our implementation. At the following code
snippet, we code the model at \pkg{ELFI}:

\begin{Code}
import elfi
import scipy.stats as ss
import numpy as np

def simulator(t1, batch_size = 1, random_state = None):
    c = 0.5 - 0.5**4
    if t1 < -0.5:
        y = ss.norm(loc = -t1-c, scale = 1).rvs(random_state = random_state)
    elif t1 <= 0.5:
        y = ss.norm(loc = t1**4, scale = 1).rvs(random_state = random_state)
    else:
        y = ss.norm(loc = t1-c, scale = 1).rvs(random_state = random_state)
    return y

# observation
y = 0

# Elfi graph
t1 = elfi.Prior('uniform', -2.5, 5)
sim = elfi.Simulator(simulator, t1, observed = y)
d = elfi.Distance('euclidean', sim)

# Initialize the ROMC inference method
bounds = [(-2.5, 2.5)] # bounds of the prior
parallelize = True # activate parallel execution
romc = elfi.ROMC(d, bounds = bounds, parallelize = parallelize)
\end{Code}

%% -- ROMC implementation --
\section{Implemented functionalities}

At this section, we analyze the functionalities of the training, the
inference and the evaluation part. Extended documentation for each
method can be found in \pkg{ELFI}'s
\href{https://elfi.readthedocs.io/en/latest/}{official
  documentation}. Finally, we describe how a user may extend ROMC with
its custom modules.

\subsection{Training part}
\label{subsec:training}

In this section, we describe the six functions of the training part:

\begin{Code}
>>> romc.solve_problems(n1, use_bo = False, optimizer_args = None)
\end{Code}

\noindent
This method (a) draws integers for setting the seed, (b) defines the
optimization problems and (c) solves them using either a
gradient-based optimizer (default choice) or Bayesian optimization
(BO), if \code{use_bo = True}. The tasks are completed sequentially,
as shown in Figure~\ref{fig:romc_overview}. The optimization problems
are defined after drawing \code{n1} integer numbers from a discrete
uniform distribution \(u_i \sim \mathcal{U}\{1, 2^{32}-1\}\), where
each integer \(u_i\) is the seed passed to \pkg{ELFI}'s random
simulator. The user can pass a \code{Dict} with custom parameters to
the optimizer through \code{optimizer_args}. For example, in the
gradient-based case, the user may pass \code{optimizer_args =
  \{"method": "L-BFGS-B", "jac": jac\}}, to select the
\code{"L-BFGS-B"} optimizer and use the callable \code{jac} to compute
the gradients in closed-form.

\begin{Code}
>>> romc.distance_hist(**kwargs)
\end{Code}

\noindent
This function helps the user decide which threshold \(\epsilon\) to
use by plotting a histogram of the distances at the optimal point
\(d_i(\thetab_i^*) : \{i = 1, 2, \ldots, n_1\}\) or \(\hat{d}_i^*\) in
case \code{use_bo = True}. The function forwards the keyword arguments
to the underlying \code{pyplot.hist()} of the \pkg{matplotlib}
package. In this way the user may customize some properties of the
histogram, e.g., the number of bins.

\begin{Code}
>>> romc.estimate_regions(eps_filter, use_surrogate = None, fit_models = False)
\end{Code}

\noindent
This method estimates the proposal regions around the optimal points,
following Algorithm~\ref{alg:region_construction}. The choice about
the distance function follows the previous optimization step; if a
gradient-based optimizer has been used, then estimating the proposal
region is based on the real distance \(d_i\). If BO is used, then the
surrogate distance \(\hat{d}\) is chosen. Setting
\code{use_surrogate=False} enforces the use of the real distance \(d\)
even after BO. Finally, the parameter \code{fit_models} selects
whether to fit local surrogate models \(\tilde{d}\) after estimating
the proposal regions.

\noindent
The training part includes three more functions. The function
\code{romc.fit_posterior(args*)} which is a syntactic sugar for
applying \code{.solve_problems()} and \code{.estimate_regions()} in a
single step. The function \code{romc.visualize_region(i)} plots the
bounding box around the optimal point of the \(i\)-th optimization
problem, when the parameter space is up to \(2D\). Finally,
\code{romc.compute_eps(quantile)} returns the appropriate distance
value \(d_{i=\kappa}^*\) where
\(\kappa = \lfloor quantile \cdot n \rfloor\) from the collection
\(\{ d_i^* \}, i = \{1, \ldots, n\}\) where \(n\) is the number
of accepted solutions. It can be used to automate the selection of the
threshold \(\epsilon\), e.g.,
\code{eps=romc.compute_eps(quantile=0.9)}.

\subsubsection*{Example - Training part}

In the following snippet, we put together the routines to code the
training part of the running example.

\begin{Code}
# Training (fitting) part
n1 = 500 # number of optimization problems
seed = 21 # seed for solving the optimization problems
use_bo = False # set to True for switching to Bayesian optimization

# Training step-by-step
romc.solve_problems(n1 = n1, seed = seed, use_bo = use_bo)
romc.theta_hist(bins = 100) # plot hist to decide which eps to use

eps = .75 # threshold for the bounding box
romc.estimate_regions(eps = eps) # build the bounding boxes

romc.visualize_region(i = 1) # for inspecting visually the bounding box

# Equivalent one-line command
# romc.fit_posterior(n1 = n1, eps = eps, use_bo = use_bo, seed = seed)
\end{Code}

\begin{figure}[ht]
  \begin{center}
    \input{latex_files/images/chapter3/example_gt.tex}
    \input{latex_files/images/chapter3/example_post.tex}
  \end{center}
  \caption[Histogram of distances at the one-dimensional example.]{Histogram of
    distances and visualization of a specific region.}
  \label{fig:running_example_romc_inference}
\end{figure}

\subsection{Inference part}
\label{subsec:inference}

In this section, we describe the four functions of the inference part:

\begin{Code}
>>> romc.sample(n2)
\end{Code}

\noindent
This is the basic functionality of the inference, drawing \(n_2\)
samples for each bounding box region, giving a total of
\(k \cdot n_2\) samples, where \(k < n_1\) is the number of the
optimal points remained after filtering. The samples are drawn from a
uniform distribution \(q_i\) defined over the corresponding bounding
box and the weight \(w_i\) is computed as in
Algorithm~\ref{alg:sampling_GB}.

\noindent
The inference part includes three more function. The function
\code{romc.compute_expectation(h)} computes the expectation
\(\E_{p(\thetab|\data)}[h(\thetab)]\) as in
Equation~\ref{eq:expectation}. The argument \code{h} can be any python
\code{Callable}. The method \code{romc.eval_unnorm_posterior(theta,
  eps_cutoff = False)} computes the unnormalized posterior
approximation as in Equation~\ref{eq:approx_posterior}. The method \\
\code{romc.eval_posterior(theta, eps_cutoff = False)} evaluates the
normalized posterior estimating the partition function
\(Z = \int p_{d,\epsilon}(\thetab|\data)d\thetab\) using Riemann's
integral approximation. The approximation is computationally feasible
only in a low-dimensional parametric space.

\subsubsection*{Example - Inference part}

In the following code snippet, we use the inference utilities to (a)
get weighted samples from the approximate posterior, (b) compute an
expectation and (c) evaluate the approximate posterior. We also use
some of \pkg{ELFI}'s built-in tools to get a summary of the obtained
samples. For \code{romc.compute_expectation()}, we demonstrate its use
to compute the samples mean and the samples variance. Finally, we
evaluate \code{romc.eval_posterior()} at multiple points to plot the
approximate posterior of
Figure~\ref{fig:running_example_romc_inference}. We observe that the
approximation is quite close to the ground-truth.

\begin{Code}
# Inference part
seed = 21
n2 = 50
romc.sample(n2 = n2, seed = seed)

# visualize region, adding the samples now
romc.visualize_region(i = 1)

# Visualize marginal (built-in ELFI tool)
weights = romc.result.weights
romc.result.plot_marginals(weights = weights, bins = 100, range = (-3,3))

# Summarize the samples (built-in ELFI tool)
romc.result.summary()
# Number of samples  : 19300
# Sample means: theta: -0.0116

# compute expectation
exp_val = romc.compute_expectation(h = lambda x: np.squeeze(x))
print("Expected value   : %.3f" % exp_val)
# Expected value: -0.012

exp_var = romc.compute_expectation(h = lambda x: np.squeeze(x)**2)
print("Expected variance: %.3f" % exp_var)
# Expected variance: 1.120

# eval unnorm posterior
romc.eval_unnorm_posterior(theta = 0)

# check eval posterior
romc.eval_posterior(theta = 0)
\end{Code}

\subsection{Evaluation part}
\label{subsec:evaluation}

The method \code{romc.compute_ess()} computes the Effective Sample
Size (ESS) as \(\frac{(\sum_i w_i)^2}{\sum_i w_i^2}\), which is a
useful quantity to measure how many samples actually contribute to an
expectation. For example, in an extreme case of a big population of
samples where only one has big weight, the ESS is much smaller than
the samples population.

\noindent
The method \code{romc.compute_divergence(gt_posterior, bounds, step,
  distance)} estimate the divergence between the ROMC approximation
and the ground truth posterior. Since the estimation is performed
using Riemann's approximation, the method can only work in low
dimensional spaces. The method can be used for evaluation in synthetic
examples where the ground truth is accessible. In a real-case
scenarios, where it is not expected to have access to the ground-truth
posterior, the user may set the approximate posterior obtained with
any other inference approach for comparing the two methods. The
argument \code{step} defines the step used in the Riemann's
approximation and the argument \code{distance} can be either
\code{"Jensen-Shannon"} or \code{"KL-divergence"}.

\begin{Code}
# Evaluation part
res = romc.compute_divergence(wrapper, distance = "Jensen-Shannon")
print("Jensen-Shannon divergence: %.3f" % res)
# Jensen-Shannon divergence: 0.035

nof_samples = len(romc.result.weights)
ess = romc.compute_ess()
print("Nof Samples: %d, ESS: %.3f" % (nof_samples, ess))
# Nof Samples: 19300, ESS: 16196.214
\end{Code}

\subsection{Extend the implementation with custom modules}
\label{subsec:extensibility}

ROMC is a generic LFI framework; it describes a sequence of steps for
approximating the posterior distribution without explicitly enforcing
the algorithms that solve each step. Although a specific algorithm for
each step is proposed by~\cite{Ikonomov2019}, in general, a user can
experiment with alternative algorithms. Considering that, we designed
the implementation to support flexibility.

We have specified four critical parts where a user may intervene using
custom solutions; (a) the gradient-based optimization, (b) the
Bayesian optimization, (c) the proposal region construction and (d)
the surrogate model fitting. Each of these parts corresponds to a
function inside the \code{romc.OptimisationProblem} class; (a)
\code{solve_gradients()}, (b) \code{solve_bo()}, (c)
\code{build_region()} and (d) \code{fit_local_surrogate()}
respectively. To replace any of these parts, the user must create a
custom class that inherits the basic \code{OptimisatioProblem} class
and overwrite the appropriate function.

To illustrate that in practice, suppose a user wants to fit Deep
Neural Networks instead of, the default, quadratic models as local
surrogates \(\tilde{d}_i\). Therefore, the user must create a new
class that inherits \code{OptimisationProblem} and overwrite the
\code{fit_local_surrogate(**kwargs)} function with one that fits
neural networks as local surrogates. We illustrate that in the
following snippet using the \code{neural_network.MLPRegressor} class
of the \pkg{scikit-learn} package.

\begin{Code}
class CustomOptim(OptimisationProblem):
    def __init__(self, **kwargs):
        super(CustomOptim, self).__init__(**kwargs)

    def fit_local_surrogate(self, **kwargs):
        nof_samples = 500
        objective = self.objective # the distance function

        # helper function
        def local_surrogate(theta, model_scikit):
            assert theta.ndim == 1
            theta = np.expand_dims(theta, 0)
            return float(model_scikit.predict(theta))

        # create local surrogate model as a function of theta
        def create_local_surrogate(model):
            return partial(local_surrogate, model_scikit = model)

        local_surrogates = []
        for i in range(len(self.regions)):
            # prepare dataset
            x = self.regions[i].sample(nof_samples)
            y = np.array([objective(ii) for ii in x])

            # train Neural Network
            mlp = MLPRegressor(hidden_layer_sizes = (10,10), solver = 'adam')
            model = Pipeline([('linear', mlp)])
            model = model.fit(x, y)
            local_surrogates.append(create_local_surrogate(model))

        self.local_surrogates = local_surrogates
        self.state["local_surrogates"] = True
\end{Code}

In a similar way, the user can replace any of the other three
functions. In each case, the custom function must change the value of
some variables that hold the state of the training phase. In the
following sections, we present which are these variables in each case.
When implementing the custom functions, the user may need two helping
classes; (a) \code{RomcOpimisationResult} that stores the result of
the optimization step and (b) \code{NDimBoundingBox} that stores the
bounding box. We present the definitions of these classes below and we
illustrate how to use them in each case in the next sections. Both
classes can be imported from the module
\code{elfi.methods.inference.romc}.

\begin{Code}
class RomcOpimisationResult:
    def __init__(self, x_min, f_min, hess_appr):
        Parameters
        ----------
        x_min: np.ndarray (D,) or float, the minimum point
        f_min: float, distance at the minimum point
        hess_appr: np.ndarray (DxD), Hessian approximation at x_min
        """

class NDimBoundingBox:
    def __init__(self, rotation, center, limits, eps_region):
        Parameters
        ----------
        rotation: np.array (D,D),  rotation matrix for the bounding box
        center: np.array (D,) center of the bounding box
        limits: np.ndarray, shape: (D,2), the limits of the bounding box
        eps_region: float, distance threshold
\end{Code}

\subsubsection*{(a) Extending the gradient-based optimization}

For using a custom gradient-based optimization method, the user must
overwrite the method \code{solve_gradients()}. The method can take any
number of keyword arguments. The objective of the optimization problem
(distance \(d_i\)) can be accessed through the variable
\code{self.objective}, which is a callable. In the following snippet,
after the comment \code{state variables}, we present the state
variables that must be set by the method. The class
\code{RomcOpimisationResult(x, y, jac, hess_inv)} must be used to
store the optimization result.

\begin{Code}
def solve_gradients(self, **kwargs):
    # inputs
    # self.objective: the distance function, Callable: np.ndarray (D,) -> float

    # custom solution procedure
    x, y, jac, hess_inv = ...
    success = True/False # whether gradient-based optimization was succesful

    # state variables
    self.state["attempted"] = True
    if success:
        self.result = RomcOpimisationResult(x, y, jac, hess_inv)
        self.state["solved"] = True
        return True
    else:
        return False
\end{Code}

\subsubsection*{(b) Extending the Bayesian Optimization}

For replacing the Bayesian Optimization function, the procedure is
similar to the gradient-based case. As presented in the following
snippet, the only added state variables are (a) setting
\code{self.surrogate = custom_surrogate}, where
\code{custom_surrogate} is a \code{Callable} with signature
\code{np.ndarray (D,) -> float}, and setting
\code{self.state["has_fit_surrogate"] = True} if the optimization is
successful.

\begin{Code}
def solve_bo(self, **kwargs):
    # inputs
    # self.objective: the distance function, Callable: np.ndarray (D,) -> float

    # custom procedure
    x, y, jac, hess_inv = ...
    custom_surrogate = ...
    success = True/False # whether Bayesian optimization was succesful

    # state variables
    self.state["attempted"] = True
    if success:
        self.result = RomcOpimisationResult(x, y, jac, hess_inv)
        self.surrogate = custom_surrogate
        self.state["solved"] = True
        self.state["has_fit_surrogate"] = True
        return True
    else:
        return False
\end{Code}

\subsubsection*{(c) Extending the proposal region construction}

For replacing the proposal region construction the user must overwrite
the \code{build_region} method. Using the objective function \(d_i\)
(\code{self.objective}) the method must estimate a list of bounding
boxes as a \code{List} with \code{NDimBoundingBox} instances and set
the state variables presented below:

\begin{Code}
def build_region(self, **kwargs):
    # inputs
    # self.objective: the distance function, Callable: np.ndarray (D,) -> float

    # custom build_region method
    bounding_box: List[NDimBoundingBox] = ...
    success = True/False # whether region built correctly

    # state variables
    self.eps_region = eps_region
    if success:
        # construct region
        self.regions = bounding_box
        self.state["region"] = True
        return True
    else:
        return False
\end{Code}

\subsubsection*{(d) Extending the surrogate model fitting}

For replacing the surrogate model fitting the user must overwrite the
\code{fit_local_surrogate} method. Using the objective function
\(d_i\) (\code{self.objective}) and the estimated bounding boxes
(\code{self.regions}), the method must estimate a list of local
surrogates, one for each region, as a \code{List} with callables and
set the state variables presented below:

\begin{Code}
def fit_local_surrogate(self, **kwargs):
    # inputs
    # self.objective: Callable np.ndarray (D,) -> float, the distance function
    # self.regions: List[NDimBoundingBox], the bounding boxes

    # custom local surrogates
    local_surrogates: List[Callable] = ...
    success = True/False # whether local surrogates fit correctly

    # side-effects
    if success:
        self.local_surrogates = custom_surrogates
        self.state["local_surrogates"] = True
        return True
    else:
        return False
\end{Code}
